Multiway Spherical Clustering via \\ Degree-Corrected Tensor Block Models

### 2022-01-17 tensor_clustering/tensor_clustering_ieee/ieee_v1.tex
update

{\bf Parameter Selection.} Note that we assume the number of clusters $r$ is given for the algorithm. We impose the Bayesian information criterion (BIC)
and choose the number of cluster that minimizes BIC, i.e., under dTBM~\eqref{eq:model_tensor},
\begin{equation}
    \hat r = \argmin_{r \in \mathbb{Z}_+ } \of{ -p^K \tL_{\tY}(\hat z(r), \hat \tS(r), \hat \mtheta(r)) + p_e(r) K \log p},
\end{equation}
where the triplet $(\hat z(r), \hat \tS(r), \hat \mtheta(r))$ are estimated parameters with given number of cluster $r$, $\tL_{\tY}$ denote the log-likelihood with observations $\tY$, and $p_e = r^K + p(\log r + 1) - 1$ is the effective number of parameters. Particularly, we obtain the estimate $\hat \mtheta(r)$ by first calculating the reduced tensor $\tY^{\text{d}}$ with $\hat z(r)$ and then normalizing the row norms $\onormSize{}{\mY_{i:}^{\text{d}}}$ to 1 in each cluster; i.e., 
\begin{equation}
    \hat \theta(i,r) = \frac{\onormSize{}{\hat \mY^{\text{d}}(r)_{i:}}}{\sum_{j: \hat z(j,r) =  \hat z(i,r)} \onormSize{}{\hat \mY^{\text{d}}(r)_{j:}}},
\end{equation}
where $\hat \mY^{\text{d}}(r) = \mat(\hat \tY^{\text{d}}(r))$ and 
\begin{equation}
    \hat \tY^{\text{d}}(r)(i,a_2,\ldots,a_K) = 
\text{Ave}\{\tY(i,i_2,\ldots,i_K): \hat z(i_k, r) = a_k, k \neq 1 \}.
\end{equation}

@@

\begin{table}[hbt]
\color{blue}
    \centering
    \begin{tabularx}{\textwidth}{c *{12}{Y}}
    \toprule
    Settings & \multicolumn{3}{c}{$p = 50, \sigma^2 = 0.25$} & \multicolumn{3}{c}{$p = 50, \sigma^2 = 1$} & \multicolumn{3}{c}{$p = 80, \sigma^2 = 0.25$} & \multicolumn{3}{c}{$p = 80, \sigma^2 = 1$}\\
    \cmidrule(lr){2-4} \cmidrule(l){5-7} \cmidrule(l){8-10} \cmidrule(l){11-13}
         True number of clusters $r$ & 2&3 & 4 & 2 & 3 & 4 & 2 & 3 & 4 & 2 & 3 & 4  \\
         \midrule
         Estimated number of clusters $\hat r$ &  2 & 3 & 3.9& 2 & 3 & 3.1 & 2 & 3 & 4 & 2 & 3& 3.9   \\
     \bottomrule
    \end{tabularx}
    \caption{Estimated number of clusters given by BIC criterion under the small noise $(\sigma^2 = 0.25)$ and large noise $(\sigma^2 = 0.5)$ cases. We report the averaged estimated number of clusters $\hat r$ over 30 duplication.}
    \label{tab:select}
\end{table}

The third experiment evaluates the empirical performance of the BIC criterion to select unknown number of cluster. We generate the data from an order-3 Gaussian model with $p = \{50,80\}$, $r = \{2,4\}$ and consider the noise $\sigma^2 \in \{ 0.25,1\}$ with $\alpha = 400$.  Table~\ref{tab:select} implies that our BIC criterion exactly choose the true $r$ under all the settings with small $r = 2, 3$.  Note that the BIC underestimates the large true number of clusters $(r = 4)$ with smaller SNR $(\sigma^2 = 1)$, and the accuracy immediately increases with larger dimension $p = 80$. The improvement follows by the fact that a larger dimension $p$ indicates a larger sample size in the block model. Therefore, we conclude that BIC criterion is a reasonable way to tune the number of clusters.
}
% {
% \color{red} 
% More experiments:
% \begin{enumerate}
%     \item Initialization impact to Angle-based iteration.(Figure 4 in Han)
%     \item Imbalanced cluster size.(Figure 9 in Han)
%     \item Non-degree comparison. 
% \end{enumerate}
% }

@@

    {\bf Step 3 (Derivation of contraction inequality):} In this step we derive the upper bound of $\zeta_{ib}$ and thereof obtain the contraction inequality. Note that 
    \begin{align}
        \zeta_{ib}^{(t)} &= \theta(i) \onormSize{}{ [ \mS_{ z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2 \ind \offf{z^{(t+1)}(i) = b, \frac{1}{2}\onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2 \leq (\theta(i) m)^{-1} F_{ib}^{(t)} + G_{ib}^{(t)} + H_{ib}^{(t)} }\\
        & \leq \theta(i) \onormSize{}{ [ \mS_{ z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2 \ind \offf{z^{(t+1)}(i) = b, \frac{1}{4}\onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2 \leq (\theta(i) m)^{-1} F_{ib}^{(t)} + G_{ib}^{(t)} } \\
        & \lesssim \ind \offf{ z^{(t+1)}(i) = b} \of{  \frac{(F_{ib}^{(t)})^2}{ \theta(i)m^2\onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2} + \frac{\theta(i)(G_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2}    }\\
        &\leq C \ind \offf{ z^{(t+1)}(i) = b} \of{  \frac{(F_{ib}^{(t)})^2}{ \theta(i)m^2\onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2} + \frac{\theta(i)(G_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2}    },\label{eq:zeta}
    \end{align}
    where the first inequality follows by the inequality~\eqref{eq:hib} in Lemma~\ref{lem:upper_fgh}, and the last inequality follows by the assumption that $\min_{i \in [p]} \theta(i) \geq c$. Following \citet[Step 4, Proof of Theorem 2]{han2020exact} and Lemma~\ref{lem:upper_fgh}, we have 
    \begin{align}\label{eq:f_sum}
          \frac{1}{p} \sum_{i \in [p]}\sum_{b \in [r]/z(i)} \ind \offf{ z^{(t+1)}(i) = b} \frac{(F_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2} \leq C_1 L^{(t)},
    \end{align}
    and 
    \begin{align}
         \frac{1}{p} \sum_{i \in [p]}\sum_{b \in [r]/z(i)} \ind \offf{ z^{(t+1)}(i) = b} \frac{\theta(i)(G_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2}  &\lesssim \frac{1}{p} \sum_{i \in [p]}\theta(i) \sum_{b \in [r]/z(i)}   \ind \offf{ z^{(t+1)}(i) = b}  (\Delta_{\min}^2 + L^{(t)}) \\
         & \leq C_2 (L^{(t+1)} + L^{(t)}), \label{eq:g_sum}
    \end{align}
    where the last inequality follows by the definition of $L^{(t)}$ and the constraint of $\mtheta$ in parameter space~\eqref{eq:family}. Choosing the constants $C, C_1, C_2$ in~\eqref{eq:zeta}~\eqref{eq:f_sum}~\eqref{eq:g_sum} carefully, we have 
    \begin{equation}\label{eq:zeta_upper}
        \frac{1}{p}\sum_{i \in [p] }  \sum_{b \in [r]/z(i)}  \zeta_{ib}^{(t)} \leq  C_4 L^{(t+1)}  + C_5 L^{(t)},
    \end{equation}
    for some small positive constants $C_4, C_5$. Plugging the inequality~\eqref{eq:zeta_upper} into the decomposition~\eqref{eq:decomp2}, we obtain the contraction inequality 
    \begin{equation}\label{eq:decomp_t}
          L^{(t+1)} \leq M \xi  + \rho L^{(t)},
    \end{equation}
     where $M \geq 1$ and $\rho \in [0,1)$ is the contraction parameter.

>>>>

{\bf Parameter Selection.} Note that we assume the number of clusters $r$ is given for the algorithm. We impose the Bayesian information criterion (BIC)
and choose the number of cluster that minimizes BIC, i.e., under dTBM~\eqref{eq:model_tensor},
\begin{equation}
    \hat r = \argmin_{r \in \mathbb{Z}_+ } \of{ -p^K \tL_{\tY}(\hat z(r), \hat \tS(r), \hat \mtheta(r)) + p_e(r) K \log p},
\end{equation}
where the triplet $(\hat z(r), \hat \tS(r), \hat \mtheta(r))$ are estimated parameters with given number of cluster $r$, $\tL_{\tY}$ denote the log-likelihood with observations $\tY$, and $p_e = r^K + p(\log r + 1) - r$ is the effective number of parameters. Particularly, we obtain the estimate $\hat \mtheta(r)$ by first calculating the reduced tensor $\tY^{\text{d}}$ with $\hat z(r)$ and then normalizing the row norms $\onormSize{}{\mY_{i:}^{\text{d}}}$ to 1 in each cluster; i.e., 
\begin{equation}
    \hat \theta(i,r) = \frac{\onormSize{}{\hat \mY^{\text{d}}(r)_{i:}}}{\sum_{j: \hat z(j,r) =  \hat z(i,r)} \onormSize{}{\hat \mY^{\text{d}}(r)_{j:}}},
\end{equation}
where $\hat \mY^{\text{d}}(r) = \mat(\hat \tY^{\text{d}}(r))$ and 
\begin{equation}
    \hat \tY^{\text{d}}(r)(i,a_2,\ldots,a_K) = 
\text{Ave}\{\tY(i,i_2,\ldots,i_K): \hat z(i_k, r) = a_k, k \neq 1 \}.
\end{equation}

@@

% \begin{table}[hbt]
% \color{blue}
%     \centering

%     \begin{tabularx}{\textwidth}{c *{12}{Y}}
%     \toprule
%     Settings & \multicolumn{3}{c}{$p = 50, \sigma^2 = 0.25$} & \multicolumn{3}{c}{$p = 50, \sigma^2 = 1$} & \multicolumn{3}{c}{$p = 80, \sigma^2 = 0.25$} & \multicolumn{3}{c}{$p = 80, \sigma^2 = 1$}\\
%     \cmidrule(lr){2-4} \cmidrule(l){5-7} \cmidrule(l){8-10} \cmidrule(l){11-13}
%          True number of clusters $r$ & 2&3 & 4 & 2 & 3 & 4 & 2 & 3 & 4 & 2 & 3 & 4  \\
%          \midrule
%          Estimated number of clusters $\hat r$ &  2 & 3 & 3.9(0.3)& 2 & 3  & 3.1(0.5) & 2  & 3  & 4  & 2  & 3 & 3.9(0.31)   \\
%      \bottomrule
%     \end{tabularx}
%     \caption{Estimated number of clusters given by BIC criterion under the small noise $(\sigma^2 = 0.25)$ and large noise $(\sigma^2 = 0.5)$ cases. Numbers in parentheses are standard deviations of $\hat r$ over 30 replications.}
%     \label{tab:select}
% \end{table}

\begin{table}[hbt]
\color{blue}
    \centering

    \begin{tabularx}{\textwidth}{c *{8}{Y}}
    \toprule
    Settings & \multicolumn{2}{c}{$p = 50, \sigma^2 = 0.25$} & \multicolumn{2}{c}{$p = 50, \sigma^2 = 1$} & \multicolumn{2}{c}{$p = 80, \sigma^2 = 0.25$} & \multicolumn{2}{c}{$p = 80, \sigma^2 = 1$}\\
    \cmidrule(lr){2-3} \cmidrule(l){4-5} \cmidrule(l){6-7} \cmidrule(l){8-9}
         True number of clusters $r$ & 2 & 4 & 2  & 4 & 2  & 4 & 2 & 4  \\
         \midrule
         Estimated number of clusters $\hat r$ &  2(0)  & 3.9(0.25)& 2(0)    & 3.1(0.52) & 2(0)    & 4(0)   & 2(0)    & 3.9(0.31)   \\
     \bottomrule
    \end{tabularx}
    \caption{Estimated number of clusters given by BIC criterion under the small noise $(\sigma^2 = 0.25)$ and large noise $(\sigma^2 = 0.5)$ cases. Numbers in parentheses are standard deviations of $\hat r$ over 30 replications.}
    \label{tab:select}
\end{table}

The third experiment evaluates the empirical performance of the BIC criterion to select unknown number of cluster. We generate the data from an order-3 Gaussian model with $p = \{50,80\}$, $r = \{2,4\}$ and consider the noise $\sigma^2 \in \{ 0.25,1\}$ with $\alpha = 400$.  Table~\ref{tab:select} implies that our BIC criterion exactly choose the true $r$ under all the settings with small $r = 2$.  Note that the BIC underestimates the large true number of clusters $(r = 4)$ with smaller SNR $(\sigma^2 = 1)$, and the accuracy immediately increases with larger dimension $p = 80$. The improvement follows by the fact that a larger dimension $p$ indicates a larger sample size in the block model. Therefore, we conclude that BIC criterion is a reasonable way to tune the number of clusters.
}
% {
% \color{red} 
% More experiments:
% \begin{enumerate}
%     \item Initialization impact to Angle-based iteration.(Figure 4 in Han)
%     \item Imbalanced cluster size.(Figure 9 in Han)
%     \item Non-degree comparison. 
% \end{enumerate}
% }

@@

    {\bf Step 3 (Derivation of contraction inequality):} In this step we derive the upper bound of $\zeta_{ib}$ and thereof obtain the contraction inequality. Note that 
    \begin{align}
        \zeta_{ib}^{(t)} &= \theta(i) \onormSize{}{ [ \mS_{ z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2 \ind \offf{z^{(t+1)}(i) = b, \frac{1}{2}\onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2 \leq (\theta(i) m)^{-1} F_{ib}^{(t)} + G_{ib}^{(t)} + H_{ib}^{(t)} }\\
        & \leq \theta(i) \onormSize{}{ [ \mS_{ z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2 \ind \offf{z^{(t+1)}(i) = b, \frac{1}{4}\onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2 \leq (\theta(i) m)^{-1} F_{ib}^{(t)} + G_{ib}^{(t)} } \\
        & \lesssim \ind \offf{ z^{(t+1)}(i) = b} \of{  \frac{(F_{ib}^{(t)})^2}{ \theta(i)m^2\onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2} + \frac{\theta(i)(G_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2}    }\\
        &\leq C \ind \offf{ z^{(t+1)}(i) = b} \of{  \frac{(F_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2} + \frac{\theta(i)(G_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2}    },\label{eq:zeta}
    \end{align}
    where the first inequality follows by the inequality~\eqref{eq:hib} in Lemma~\ref{lem:upper_fgh}, and the last inequality follows by the assumption that $\min_{i \in [p]} \theta(i) \geq c$. Following \citet[Step 4, Proof of Theorem 2]{han2020exact} and Lemma~\ref{lem:upper_fgh}, we have 
    \begin{align}\label{eq:f_sum}
          \frac{1}{p} \sum_{i \in [p]}\sum_{b \in [r]/z(i)} \ind \offf{ z^{(t+1)}(i) = b} \frac{(F_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2} \leq C_1 L^{(t)},
    \end{align}
    and 
    \begin{align}
         \frac{1}{p} \sum_{i \in [p]}\sum_{b \in [r]/z(i)} \ind \offf{ z^{(t+1)}(i) = b} \frac{\theta(i)(G_{ib}^{(t)})^2}{ \onormSize{}{ [ \mS_{z(i):}  ]^s - [ \mS_{b:}  ]^s  }^2}  &\lesssim \frac{1}{p} \sum_{i \in [p]}\theta(i) \sum_{b \in [r]/z(i)}   \ind \offf{ z^{(t+1)}(i) = b}  (\Delta_{\min}^2 + L^{(t)}) \\
         & \leq C_2 (L^{(t+1)} + L^{(t)}), \label{eq:g_sum}
    \end{align}
    where the last inequality follows by the definition of $L^{(t)}$ and the constraint of $\mtheta$ in parameter space~\eqref{eq:family}. Choosing the constants $C, C_1, C_2$ in~\eqref{eq:zeta}~\eqref{eq:f_sum}~\eqref{eq:g_sum} carefully, we have 
    \begin{equation}\label{eq:zeta_upper}
        \frac{1}{p}\sum_{i \in [p] }  \sum_{b \in [r]/z(i)}  \zeta_{ib}^{(t)} \leq  C_4 L^{(t+1)}  + C_5 L^{(t)},
    \end{equation}
    for some small positive constants $C_4, C_5$. Plugging the inequality~\eqref{eq:zeta_upper} into the decomposition~\eqref{eq:decomp2}, we obtain the contraction inequality 
    \begin{equation}\label{eq:decomp_t}
          L^{(t+1)} \leq M \xi  + \rho L^{(t)},
    \end{equation}
     where $M \geq 1$ and $\rho \in [0,1)$ is the contraction parameter.
